---
title: "Muscat testing"
author: "Kristen Wells"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    highlight: "tango"
    df_print: "paged"
    code_folding: "hide"
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = F, 
  warning = F,
  comment = ""
)
library(here)
knitr::opts_knit$set(root.dir = here())
```

# Background
This document tests muscat using different parameters to ensure I am running it correctly

```{r load-packages}
library(here)
library(scAnalysisR)
library(pheatmap)
library(tidyverse)
library(muscat)
library(SingleCellExperiment)
library(Seurat)
library(DESeq2)
library(cowplot)

source(here("src/scripts/muscat_plotting_functions.R"))
```

```{r load-data}
all_colors <- readRDS(file = file.path("files/all_colors.rds"))


final_colors <- all_colors$cell_type_colors

final_colors <- final_colors[names(final_colors) != "Translational_Intermediate"]

tetramer_colors <- all_colors$tetramer_colors

sample_colors <- all_colors$sample_colors


status_colors <- all_colors$status_colors

normalization_method <- "log" # can be SCT or log
# Set theme
ggplot2::theme_set(ggplot2::theme_classic(base_size = 10))

normalization_method <- "log" # can be SCT or log

#args <- commandArgs(trailingOnly = TRUE)

args <- c("merged", here("results"), "", here("files/sample_info.tsv"))

sample <- args[[1]]
sample <- gsub("__.*", "", sample)
#sample <- "merged"

results_dir <- args[[2]]
#results_dir <- here("results")

sample_info <- args[[4]]
#sample_info <- here("files/sample_info.tsv")

sample_info <- read.table(sample_info, fill = TRUE, header = TRUE)

sample_info <- sample_info[sample_info$sample == sample,]
vars.to.regress <- NULL
HTO <- sample_info$HTO
ADT <- sample_info$ADT
hash_ident <- sample_info$hash_ident

ADT_pca <- sample_info$adt_pca
run_adt_umap <- sample_info$adt_umap

RNA_pcs <- sample_info$PCs
resolution <- sample_info$resolution
batch_correction <- sample_info$batch_correction

if(normalization_method == "SCT"){
  SCT <- TRUE
  seurat_assay <- "SCT"
} else {
  SCT <- FALSE
  seurat_assay <- "RNA"
}

# Set directories
save_dir <- file.path(results_dir, "R_analysis", sample)

# Read in data
seurat_data <- readRDS(file.path(save_dir, "rda_obj", "seurat_processed_no_doublet.rds"))
DefaultAssay(seurat_data) <- "RNA"

seurat_data <- DietSeurat(object = seurat_data, assays = "RNA")
```

# Testing batch correction
Below I set up the data and test batch correction

```{r muscat-setup}
# All B cells ------------------------------------------------------------------

## Rename columns for compatability --------------------------------------------

test_mapping <- c("DNA.tet" = "Other",
                  "GAD.tet" = "Islet_Reactive",
                  "IA2.tet" = "Islet_Reactive",
                  "INS.tet" = "Islet_Reactive",
                  "Islet_Multi_Reactive" = "Islet_Reactive",
                  "Negative" = "Other",
                  "Other_Multi_Reactive" = "Other",
                  "TET.tet" = "Other")

seurat_data$test_id <- test_mapping[seurat_data$tet_name_cutoff]

# Name columns for muscat sample_id, group_id, cluster_id
seurat_data$sample_id <- paste(seurat_data$sample, seurat_data$test_id,
                               sep = "_")
seurat_data$group_id <- seurat_data$Status

seurat_data$celltype <- "bcells"
seurat_data$cluster_id <- seurat_data$celltype


# Change to single cell experiment
sce <- as.SingleCellExperiment(seurat_data)
sce <- prepSCE(sce,
               kid = "cluster_id",
               gid = "group_id",
               sid = "sample_id",
               drop = FALSE)

# Aggregate data to pseudotbulk using default parameters from tutorial
pb <- aggregateData(sce, assay = "counts",
                    fun = "sum",
                    by = c("cluster_id", "sample_id"))


pb_mds <- pbMDS(pb)
pb_mds

### Check for batch ------------------------------------------------------------
plot_data <- pb_mds$data

# fix the dates
colData(pb)$Collection_Date <- zoo::as.Date(colData(pb)$Collection.Date)
colData(pb)$Collection_Date <- gsub("-", "_", colData(pb)$Collection_Date)
colData(pb)$processed_date <- zoo::as.Date(as.numeric(colData(pb)$date.processed.for.scSeq))
colData(pb)$processed_date <- gsub("-", "_", colData(pb)$processed_date)


colData(pb)$Collection_Date <- ifelse(colData(pb)$Collection_Date == 
                                        "2022_02_08", "2023_02_08",
                                      colData(pb)$Collection_Date)


### Run batch correction -------------------------------------------------------

# Definitely grouping by individual, somewhat by family, can't correct by
pb$test_full_new_status <- paste(pb$Status, pb$test_id,
                                          sep = "_")
dds <- DESeqDataSet(pb, design = ~processed_date + test_full_new_status)

vsd_one <- vst(dds, blind = F)

mat <- assay(vsd_one)

# Day is the batch, the model.matrix should include everything but the batch
mat <- limma::removeBatchEffect(mat, vsd_one$processed_date,
                                design = model.matrix(~vsd_one$test_full_new_status))

vsd_two <- vsd_one
assay(vsd_two) <- mat

mod_mat <- model.matrix(design(dds), colData(dds))

```

Below are plots testing batch correction
A-B) Status before and after correction
C-D) Processed date before and after correction
E-F) Sample before and after correction
G-H) Tetramer call before and after correciton

```{r plots, fig.height = 16, fig.width = 10}

p1 <- plot_pca(vsd_one, group_by = "Status") +
        ggplot2::ggtitle("uncorrected")

p2 <- plot_pca(vsd_two, group_by = "Status") +
        ggplot2::ggtitle("corrected")

p3 <- plot_pca(vsd_one, group_by = "processed_date") +
        ggplot2::ggtitle("uncorrected")

p4 <- plot_pca(vsd_two, group_by = "processed_date") +
        ggplot2::ggtitle("corrected")

vsd_one$ID <- factor(vsd_one$ID)
vsd_two$ID <- factor(vsd_two$ID)

p5 <- plot_pca(vsd_one, group_by = "ID",
               color_palette = sample_colors) +
        ggplot2::ggtitle("uncorrected")

p6 <- plot_pca(vsd_two, group_by = "ID",
               color_palette = sample_colors) +
        ggplot2::ggtitle("corrected")

p7 <- plot_pca(vsd_one, group_by = "test_id") +
        ggplot2::ggtitle("uncorrected")

p8 <- plot_pca(vsd_two, group_by = "test_id") +
        ggplot2::ggtitle("corrected")

cowplot::plot_grid(p1, p2, p3, p4, p5, p6, p7, p8,
                   nrow = 4, ncol = 2,
                   labels = c("A", "B", "C", "D", "E", "F", "G", "H"))
```


# Muscat with correction by processed date
Below I run muscat with a correction on processed date

```{r muscat1}
### Run deseq with muscat and the new batch term -------------------------------

formula <- ~processed_date + test_full_new_status
cd <- as.data.frame(colData(pb))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])
diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"), ])
other <- colMeans(design[which(cd$test_id == "Other"), ])
ND_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                             & cd$group_id == "ND") , ])
ND_other <- colMeans(design[which(cd$test_id == "Other"
                                  & cd$group_id == "ND"), ])

NO_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                             & cd$group_id == "T1D") , ])
NO_other <- colMeans(design[which(cd$test_id == "Other"
                                  & cd$group_id == "T1D"), ])

AAB_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                              & cd$group_id == "AAB") , ])
AAB_other <- colMeans(design[which(cd$test_id == "Other"
                                   & cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      diabetes_antigen - other,
                                      NO_diabetes_antigen - ND_diabetes_antigen,
                                      NO_other - ND_other,
                                      AAB_diabetes_antigen - ND_diabetes_antigen,
                                      AAB_other - ND_other,
                                      NO_diabetes_antigen - AAB_diabetes_antigen,
                                      NO_other - AAB_other,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

processed_df <- data.frame(table(all_res$contrast))
colnames(processed_df) <- c("contrast", "all_processed_date_freq")

print(processed_df)

```

# Muscat with no correction
Next I ran muscat without any correction

```{r}
formula <- ~0 + test_full_new_status
cd <- as.data.frame(colData(pb))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])
diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"), ])
other <- colMeans(design[which(cd$test_id == "Other"), ])
ND_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                             & cd$group_id == "ND") , ])
ND_other <- colMeans(design[which(cd$test_id == "Other"
                                  & cd$group_id == "ND"), ])

NO_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                             & cd$group_id == "T1D") , ])
NO_other <- colMeans(design[which(cd$test_id == "Other"
                                  & cd$group_id == "T1D"), ])

AAB_diabetes_antigen <- colMeans(design[which(cd$test_id == "Islet_Reactive"
                                              & cd$group_id == "AAB") , ])
AAB_other <- colMeans(design[which(cd$test_id == "Other"
                                   & cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      diabetes_antigen - other,
                                      NO_diabetes_antigen - ND_diabetes_antigen,
                                      NO_other - ND_other,
                                      AAB_diabetes_antigen - ND_diabetes_antigen,
                                      AAB_other - ND_other,
                                      NO_diabetes_antigen - AAB_diabetes_antigen,
                                      NO_other - AAB_other,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

no_correction_df <- data.frame(table(all_res$contrast))
colnames(no_correction_df) <- c("contrast", "all_no_correction_freq")

print(no_correction_df)

```

# Muscat on tetramer alone

I wanted to next make sure this wasn't a problem with the way I was calling the function. Below I repeat on only the islet reactive cells

```{r}
pb2 <- pb[ , pb$test_id == "Islet_Reactive"]
formula <- ~0 + Status
cd <- as.data.frame(colData(pb2))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb2, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

islet_only_df <- data.frame(table(all_res$contrast))
colnames(islet_only_df) <- c("contrast", "islet_only_df_freq")

print(islet_only_df)
```

# Muscat on tetramer alone corrected

I wanted to next make sure this wasn't a problem with the way I was calling the function. Below I repeat on only the islet reactive cells with the correction

```{r}
pb2 <- pb[ , pb$test_id == "Islet_Reactive"]
formula <- ~processed_date + Status
cd <- as.data.frame(colData(pb2))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb2, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

islet_only_df_corrected <- data.frame(table(all_res$contrast))
colnames(islet_only_df_corrected) <- c("contrast", "islet_only_df_corrected_freq")

print(islet_only_df_corrected)
```

# Muscat ignoring tetramer
Because the values were so much higher for the tests done without the tetramers, I began to worry that including the tetramers here were just acting as extra data points.

```{r}
# Name columns for muscat sample_id, group_id, cluster_id
seurat_data$sample_id <- seurat_data$sample
seurat_data$group_id <- seurat_data$Status

seurat_data$celltype <- "bcells"
seurat_data$cluster_id <- seurat_data$celltype


# Change to single cell experiment
sce <- as.SingleCellExperiment(seurat_data)
sce <- prepSCE(sce,
               kid = "cluster_id",
               gid = "group_id",
               sid = "sample_id",
               drop = FALSE)

# Aggregate data to pseudotbulk using default parameters from tutorial
pb <- aggregateData(sce, assay = "counts",
                    fun = "sum",
                    by = c("cluster_id", "sample_id"))

colData(pb)$processed_date <- zoo::as.Date(as.numeric(colData(pb)$date.processed.for.scSeq))
colData(pb)$processed_date <- gsub("-", "_", colData(pb)$processed_date)


### Run deseq with muscat and the new batch term -------------------------------

formula <- ~0 + Status
cd <- as.data.frame(colData(pb))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

status_only_df <- data.frame(table(all_res$contrast))
colnames(status_only_df) <- c("contrast", "status_only_df_freq")

print(status_only_df)

```

# Repeating with correction

```{r}

formula <- ~processed_date + Status
cd <- as.data.frame(colData(pb))
design <- model.matrix(formula, cd)

ND <- colMeans(design[which(cd$group_id == "ND"), ])
NO <- colMeans(design[which(cd$group_id == "T1D"), ])
AAB <- colMeans(design[which(cd$group_id == "AAB"), ])

all_contrasts <- limma::makeContrasts(NO - ND,
                                      AAB - ND,
                                      NO - AAB,
                                      levels = design)

de_genes_all <- muscat::pbDS(pb, method = "DESeq2",
                             design = design, contrast = all_contrasts)

all_res <- lapply(de_genes_all$table, function(x){
  return(x$bcells)
})

all_res <- do.call(rbind, all_res)

all_res <- all_res %>%
  dplyr::filter(p_adj.glb < 0.05)

status_only_processed_df <- data.frame(table(all_res$contrast))
colnames(status_only_processed_df) <- c("contrast", "corrected_status_only_df_freq")

print(status_only_processed_df)
```

# Combining everything

```{r}
all_dfs <- merge(processed_df, no_correction_df, by = "contrast", all = TRUE)
all_dfs <- merge(all_dfs, islet_only_df, by = "contrast", all = TRUE)
all_dfs <- merge(all_dfs, islet_only_df_corrected, by = "contrast",
                 all = TRUE)
all_dfs <- merge(all_dfs, status_only_df, by = "contrast", all = TRUE)
all_dfs <- merge(all_dfs, status_only_processed_df, by = "contrast",
                 all = TRUE)
print(all_dfs)
```